Introducing Statistical Design of Experiments to SPARQL Endpoint Evaluation

### 2013-07-25 exp1/iswc2013.tex
Final proofreading
git-svn-id: https://svn.ifi.uio.no/repos/users/kjekje-phd@634 43ef1190-90fd-4dd7-a769-d0121f02b80e

We have chosen 8 factors each having 2 levels, ``TripleC'' the number
of triples in the data-set, 1 or 2 MTriples, ``Machine'', which is the
software and hardware platform, one larger with slower
disks\footnote{Running GNU/Linux Debian Squeeze, has 16 GB RAM, an
  Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
  configuration} and one smaller with faster disk\footnote{Running
  Debian Wheezy, has 8 GB RAM, an Intel Core i7-3520M CPU and a single
  SSD on SATA-III}.

@@

\begin{center}\small
  \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.53\textwidth}|}
    \hline
    \textbf{``Union''} & \textbf{``Lang''} \\[-2ex]
    \begin{verbatim}
{
  ?o1 dbpprop:longd ?long ;
      dbpprop:latd  ?lat .
} UNION {
  ?o1 geo:long ?long ;
      geo:lat ?lat .
}\end{verbatim} 
&
\begin{verbatim}
FILTER langMatches( lang(?l1), "fr" ) 
\end{verbatim} 
\\    \hline
    \textbf{``Range''} & \textbf{``Optional''} \\[-2ex]
    \begin{verbatim}
FILTER (?l2 > 800000)\end{verbatim} 
 & 
    \begin{verbatim}
OPTIONAL { ?o1 foaf:homepage ?o6 . }\end{verbatim} 
\\    \hline
  \end{tabular}
\end{center}
These fragments has been carefully designed for illustrative utility,
as well as suitable selectivity. They are not themselves important,
they only serve to illustrate certain examples of possible
factors. When the experiment is run, they are combined to yield a
query, which is then sent to a SPARQL endpoint. The SPARQL endpoint is
itself set up using 4store (see \cite{harris20094store}) version
1.1.5.

@@

To proceed beyond a visual judgment of significance, we may use
Lenth's method\cite{lenthmethod}, which is a simple method for
estimating significance in an experiment with no replicates. In
Table~\ref{tab:effects}, the most important effects are listed.  In
Figure~\ref{fig:fullnormal}, Lenth's method is used to label the
significant effects. We see that we have many very significant effects
(unfortunately so many that some labels are illegible). We also note that
``Implement'' itself is significantly negative. We also note that the
``Implement:Lang'' interaction is significantly positive, as expected
from our setup, where ``Lang'' was the detrimental side-effect we
simulated.

@@

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. It is of critical
importance that we understand why some effects are significant. That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. By carefully inspecting Table~\ref{tab:pvaluesfull},
we see that whenever ``TripleC'' is at the low level, the
new implementation is always best, only at the higher level, the old
implementation may be better. This is a clear indication that the
separation between levels for the sizes of the data-set is too small
and should be investigated further. Such reasoning should be applied
to all significant effects.

@@

We lose explanatory power due to \emph{aliasing}, for example, the
interaction labelled ``TripleC:BGPComp'' may be aliased with the
``Machine:Range'' interaction. That is to say, a detected increase in
run time from larger Basic Graph Patterns for large databases, can
also be explained by a less powerful machine that is worse at
evaluating FILTER clauses with ranges. They are
indistinguishable. This may or may not cause a problem. In many cases,
we may not be interested in these effects, we may only be interested
in ``Implement'' and its interactions. Moreover, it is possible (even
easy using the FrF2 package in R) to declare which effects must not be
aliased, and determine the size of the experiment based on that. Such
a main effect or two-factor interaction is called \emph{clear} if none
of its aliases are other main effects or two-factor
interactions. Another possibility to shrink the size of the experiment
is if we \textit{a priori} can say that some effects are
negligible. We have not found such assumptions to be tenable in our
case.

>>>>

We have chosen 8 factors each having 2 levels, ``TripleC'' the number
of triples in the data-set, 1 or 2 MTriples, ``Machine'', which is the
software and hardware platform, one larger with slower
disks\footnote{Running GNU/Linux Debian Squeeze, has 16 GB RAM, an
  Intel Core2 Duo E8500 CPU and two Parallel-ATA disks in a RAID-1
  configuration} and one smaller with a faster disk\footnote{Running
  Debian Wheezy, has 8 GB RAM, an Intel Core i7-3520M CPU and a single
  SSD on SATA-III}.

@@

\begin{center}\small
  \begin{tabular}{|@{~~}p{0.40\textwidth}|@{~~}p{0.53\textwidth}|}
    \hline
    \textbf{``Union''} & \textbf{``Lang''} \\[-2ex]
    \begin{verbatim}
{
  ?o1 dbpprop:longd ?long ;
      dbpprop:latd  ?lat .
} UNION {
  ?o1 geo:long ?long ;
      geo:lat ?lat .
}\end{verbatim} 
&
\begin{verbatim}
FILTER langMatches( lang(?l1), "fr" ) 
\end{verbatim} 
\\    \hline
    \textbf{``Range''} & \textbf{``Optional''} \\[-2ex]
    \begin{verbatim}
FILTER (?l2 > 800000)\end{verbatim} 
 & 
    \begin{verbatim}
OPTIONAL { ?o1 foaf:homepage ?o6 . }\end{verbatim} 
\\    \hline
  \end{tabular}
\end{center}
These fragments have been carefully designed for illustrative utility,
as well as suitable selectivity. They are not themselves important,
they only serve to illustrate certain examples of possible
factors. When the experiment is run, they are combined to yield a
query, which is then sent to a SPARQL endpoint. The SPARQL endpoint is
itself set up using 4store (see \cite{harris20094store}) version
1.1.5.

@@

To proceed beyond a visual judgment of significance, we may use
Lenth's method\cite{lenthmethod}, which is a simple method for
estimating significance in an experiment with no replicates. In
Table~\ref{tab:effects}, the most important effects are listed.  In
Figure~\ref{fig:fullnormal}, Lenth's method is used to label the
significant effects. We see that we have many very significant effects
(unfortunately so many that some labels are illegible). We also note that
``Implement'' itself is significantly negative and that the
``Implement:Lang'' interaction is significantly positive, as expected
from our setup, where ``Lang'' was the detrimental side-effect we
simulated.

@@

The main objective of this paper is not to establish that the
optimization is significant, it is merely a simulation. It is to
establish a critical practice of evaluations. We must therefore
understand why some effects are significant. That
``Implement'' is significant is hardly a surprise, that is what we
worked for. That the main effects ``Union'', ``Optional'' and ``Lang''
are strong is also due to that these are fairly demanding things to
evaluate. Since the delay we inserted affects all kinds of joins, it
is also intuitive that interactions between ``Implement'' and those
that require joins are strong. The strong ``BGPComp:Lang'' and
``Lang:Union'' interactions might be due to that the many more triples
needs to be searched for a language tag in one case. However, the
positive ``Implement:TripleC'' interaction evades such explanations;
in fact, it hints that our optimization may not work as well for
larger databases. By carefully inspecting Table~\ref{tab:pvaluesfull},
we see that whenever ``TripleC'' is at the low level, the
new implementation is always best, only at the higher level, the old
implementation may be better. This is a clear indication that the
separation between levels for the sizes of the data-set is too small
and should be investigated further. Such reasoning should be applied
to all significant effects.

@@

We lose explanatory power due to \emph{aliasing}; for example, the
interaction labelled ``TripleC:BGPComp'' may be aliased with the
``Machine:Range'' interaction. That is to say, a detected increase in
run time from larger Basic Graph Patterns for large databases, can
also be explained by a less powerful machine that is worse at
evaluating FILTER clauses with ranges. They are
indistinguishable. This may or may not cause a problem. In many cases,
we may not be interested in these effects, we may only be interested
in ``Implement'' and its interactions. Moreover, it is possible (even
easy using the FrF2 package in R) to declare which effects must not be
aliased, and determine the size of the experiment based on that. Such
a main effect or two-factor interaction is called \emph{clear} if none
of its aliases are other main effects or two-factor
interactions. Another possibility to shrink the size of the experiment
is if we \textit{a priori} can say that some effects are
negligible. We have not found such assumptions to be tenable in our
case.
